---
title: "Lab10: Discrete Survival Models"
output: 
  rmarkdown::html_vignette:
    keep_md: yes
vignette: >
  %\VignetteIndexEntry{Lab10_CJS}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
   warning = FALSE, message = FALSE,
   fig.align = 'center', fig.retina = 2,
  fig.height=8, fig.width=8,
  collapse = TRUE,
  comment = "#>"
)
library(dplyr)
library(knitr)
library(ggplot2)
source(here::here("R/zzz.R"))
```

```{r setup, echo=FALSE}
library(WILD8370)
library(nimble)
library(coda)
library(mcmcOutput)
library(MCMCvis)
library(terra)
```

In this exercise, we'll look at how our estimates might change when we fit data with imperfect detection to both a known-fate and a CJS model. 


## Homework
1. 


2. Using only the capture records for *Rana draytonii*, write a CJS model that models capture probability as a function of both sex and a random effect of year. Make a ggplot showing detection probability across time, with different colors for the different sexes.

3. Fix someone's bad code

4.  On a 1-10 scale, with 1 being the worst week ever and 10 being the best, how would you rate this week's content? What lingering questions/confusion about the lecture or lab do you still have?
