---
title: 'Lab 3: Basic MCMC'
author: "Clark Rushing and Heather Gaya"
output: 
  rmarkdown::html_vignette:
    keep_md: yes
vignette: >
  %\VignetteIndexEntry{Lab3_BasicMCMC}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
   warning = FALSE, message = FALSE,
   fig.align = 'center', fig.retina = 2,
  collapse = TRUE,
  comment = "#>"
)
library(dplyr)
library(knitr)
source(here::here("R/zzz.R"))
```

```{r setup, echo=FALSE}
library(WILD8370)
```

In this activity, we will use `R` to create a simple Metropolis sampler to estimate the posterior distributions of the parameters in a linear regression model. This sounds like a giant task but once you understand the mechanisms, it's not too hard.

------------------------------------------------------------------------

**Objectives**

-   Create a Metropolis sampler to estimate the parameters of a linear regression model

-   Create custom functions to avoid copying/pasting code

------------------------------------------------------------------------

# Markov chain Monte Carlo

The basic steps of the Metropolis sampler are:

0)  Choose an initial value for $\theta$ (call it $\theta^1$)

1)  Propose a new value $\theta^*$ from a proposal distribution

2)  Compute the probability of accepting $\theta^*$ using the joint distributions evaluated at $\theta^*$ and the previous value $\theta^{k-1}$

3)  Accept $\theta*$ (i.e., $\theta^{k} = \theta^*$) with the probability estimated in step 2, otherwise retain the previous value (i.e., $\theta^{k} = \theta^{k-1}$

# Why Bother?

First, let's simulate some fake data from a binomial distribution as an example. Maybe we're flipping a coin and want to know if it's fair. We flip it 20 times and these are the results we get:

```{r}
set.seed(20)
x <- rbinom(20, size = 1, prob = .2)
x
```

Let's take a look at our data with a table:

```{r}
table(x)
```

Looks like we have 16 zeros ("failures") and 4 1's ("successes"). Probably not a fair coin. But what if we want to estimate the probability of success? If it were truly fair, we'd expect a probability of 0.5.

If we reallllllly wanted to, we could try to figure this out entire "by hand" by guessing a bunch of values and using the probability density function of a binomial distribution to tell us if we were close to the correct value. Here's how that might look:

We know that the equation for a binomial distribution with size 20 is: $\frac{20!}{(20-X!)X!}(p^X)(1-p)^{20-X}$

So how do we go about finding p? Well, to start, what's the probability density of 4 successes if p = .1?

```{r}
dbinom(4, 20, .1)
```

Okay, not too high.

How about if p = .4?

```{r}
dbinom(4, 20, .4)
```

Hmm, a little lower.

What about p = .7?

```{r}
dbinom(4, 20, .7)
```

Oof, lower again.

So we could do this over and over and over again and eventually we'd see that some numbers were larger (yay) and some were smaller (less good answers). If we guessed enough numbers, we'd probably find a few that were highest and we could say that p was pretty close to whatever that value was.

Obviously picking random values of p by hand isn't very efficient. Let's have a function do it for us and graph what the result is:

```{r}
pick.ps <- function(n){
  p <- runif(n, min = 0, max = 1)
  d <- dbinom(4,20,p)
  return(data.frame(p = p, d=d))
}
```

```{r fig.width=7, fig.height=4, echo = F}
par(mfrow = c(1,2))
  plot(pick.ps(500)[,1:2], cex = .5)
 hist(pick.ps(500)[,1], breaks = seq(0,1, by = .1), main = "Tested values of p", xlab = "Value of p")
```

We can see that the value of p is probably close to .3 because that's where the highest probability density was, but we can also see that this method tried a bunch of points (AKA wastes a lot of time) over around values that we basically know aren't going to be good answers. For instance, we're pretty sure this coin isn't going to have a success probability of .9, so we would kind of prefer that the model not spend so much time testing values over there.

Is there a way we could have the function spend more time in "good" areas, trying to estimate the exact value of p and less time in "bad" areas where the answer is very unlikely? And what about for equations where the math isn't just one distribution but a bunch of distributions combined?

Why yes! It's called MCMC. And it makes our ineffective plug-and-chug methods much faster.

# Seed count model

When we generated the seed count data, we did so assuming a linear model that linked pollination visits to seed counts:

$$y_i = \alpha + \beta * x_i + \epsilon_i$$ $$\epsilon_i \sim Normal(0, \sigma^2)$$

This model includes 3 random variables that we want to estimate posterior distributions for: $\alpha$, $\beta$, and $\sigma^2$ (remember that the $y_i$ and $x_i$ are *observed* and therefore treated as *fixed* and *known* rather than random variables governed by probability distributions). So:

$$[\alpha, \beta, \sigma^2|y_i] \propto [y_i|\alpha, \beta, \sigma^2][\alpha][\beta][\sigma^2] \tag{1}$$

Sticking with the data generating model, we can define the likelihood of our model using a Normal distribution. For given values of $\alpha$, $\beta$, and $\sigma^2$, we estimate the likelihood as:

$$[y_i|\alpha, \beta, \sigma] = \prod_{i=1}^N Normal(y_i|\alpha + \beta \times x_i, \sigma^2)$$

Because $\alpha$, $\beta$, and $\sigma^2$ are random variables, we must define prior distributions for them. Because $\alpha$ and $\beta$ can be positive or negative real numbers (we'll ignore for now that we know the sign of each in our data generating model), we can use normal priors:

$$[\alpha] = Normal(\alpha|0, 50)$$

$$[\beta] = Normal(\beta|0, 50)$$

These are relatively non-informative priors for the intercept and slope coefficients.

We also need a model for $\sigma$. 

Side note: A lot of MCMC codes (such as the BUGS language) model a normal distribution as $Normal(mean, precision)$ where precision = ($1/\sigma^2$).

For $\sigma$, we will use a relatively diffuse gamma prior:

$$[\sigma] = Gamma(\sigma|1, 1)$$

# Custom functions

At each iteration of the sampler, we will need to perform the same set of tasks, namely estimating the likelihood and joint distributions for the proposed and current parameter values. We could do this by copying and pasting the code necessary for these calculations at each point in the sampler code where we need them. However, any time you find yourself copying and pasting the same code more than once or twice, it's a good idea to consider wrapping that code in a function, which will do here.

## Function for calculating likelihood

Given values of $\alpha$, $\beta$, and $\sigma$, we can estimate the likelihood in `R` using the `dnorm()` function (note that we will take the sum of the log likelihoods rather than the product of the likelihoods to avoid numerical issues):

```{r eval = FALSE}
# Calculate the predicted count for each flower
  lp <- alpha + beta * x  
  
# Calculate the likelihood of our data given the model
  sum(dnorm(y, lp, sigma, log = TRUE)) 
```

To turn this code into a function, first open a new script and call it `calc_like.R`. Save it in the `R/` sub-directory. This script will contain the function code. The function code is:

```{r}
calc_like <- function(y, alpha, beta, sigma, x) {
  lp <- alpha + beta * x
  ll <- sum(dnorm(y, lp, sigma, log = TRUE))
  return(ll)
}
```

This allows us to put in the data (`y`), covariate values (`x`), and values of $\alpha$, $\beta$, and $\sigma$ as arguments and the function will return the log likelihood.

Save this code in the script and then close the script.

## Functions for calculating prior probabilities

We also need functions to estimate the prior probabilities for specific values of $\alpha$, $\beta$, and $\sigma$. Create a new script titled `priors.R` and save it to the `R/` sub-directory. The following functions will take values of each parameter and estimate the prior probability given the prior distributions we defined above:

```{r}
priorAlpha <- function(alpha, mu = 0, sigma = 50){
  prob <- dnorm(alpha, mu, sigma, log = TRUE)
  return(prob)
} 

priorBeta <- function(beta, mu = 0, sigma = 50){
  prob <- dnorm(beta, mu, sigma, log = TRUE)
  return(prob)
} 

priorsigma <- function(sigma, shape1 = 0.1, rate1 = 0.1){
  prob <- dgamma(sigma, shape = shape1, rate = rate1, log = TRUE)
  return(prob)
} 
```

## Metropolis sampler

Now we are ready to create the sampler. Remember that when there are multiple parameters, we need to define the full conditionals. For each random variable in the model, we define the full conditional by including any element on the right hand side of eq. 1 that contains the parameter (note the change to $\sigma$):

$$[\alpha|.] = [y_i|\alpha, \beta, \sigma][\alpha]$$ $$[\beta|.] = [y_i|\alpha, \beta, \sigma][\beta]$$

$$[\sigma|.] = [y_i|\alpha, \beta, \sigma][\sigma]$$

In the sampler, we loop over each parameter, going through the Metropolis steps for each and treating the other parameters as known (based on their current value in the chain). If this in not clear, walking through the `R` code will hopefully make it more concrete (don't worry about trying to copy and run the code in this section, which is simply for illustration. In the next section, we will *slightly* modify this code to run the sampler for our seed count data).

First, we set the length of our chains, set the tuning parameter for the proposal distribution, and create an empty data frame to store the samples and a binary variable for estimating the acceptance rate:

```{r eval = FALSE}
## Length of chains
  nIter <- 10000

## Tuning parameter
  tuning <- 1.5

## Empty data frame to store posterior samples of each parameter
  mcmc_df <- data.frame(x = 1:nIter,
                        alpha = numeric(nIter),
                        beta = numeric(nIter),
                        sigma = numeric(nIter),
                        accept.alpha = 0,
                        accept.beta = 0,
                        accept.sigma = 0)
```

Next, we will randomly generate initial values for each parameter. Remember that these are quite big numbers in the data generating model so we will create initial values of the same magnitude:

```{r eval = FALSE}
## Initial values
  mcmc_df$alpha[1] <-  runif(1, 200, 300)
  mcmc_df$beta[1] <- runif(1, 25, 75)
  mcmc_df$sigma[1] <- runif(1, 0, 10)
```

The final "set up" step is to estimate the likelihood of our data given these initial values. To do this, we use the `calc_like()` function we created earlier:

```{r eval = FALSE}
## Initial likelihood
likelihood <- calc_like(y = y, alpha = mcmc_df$alpha[1], 
                       beta = mcmc_df$beta[1], sigma = mcmc_df$sigma[1], 
                       x = x) 
```

Now we create the actual sampler. Below, we will create a loop that implements these steps for each iteration of the chains. Here, we will simply go through each step to make sure you understand what it's doing. We'll start by going through the Metropolis steps for $\alpha$ (though remember the order doesn't matter):

```{r eval = FALSE}
########################
#### 1. Update alpha
########################

## 1a: Generate candidate value
  cand <- rnorm(1, mcmc_df$alpha[i - 1], tuning)

## 1b: Calculate likelihood at candidate value
  cand_like <- calc_like(y = y, alpha = cand, beta = mcmc_df$beta[i-1], sigma = mcmc_df$sigma[i-1], x = x)
  
## 1c: Calculate likelihood * prior at old value and candidate value
  jointOld <- likelihood + priorAlpha(mcmc_df$alpha[i-1]) 
  jointCand <- cand_like + priorAlpha(cand) 
  
## 1d: Acceptance probability 
  R <- min(1, exp(jointCand - jointOld))

## 1e: Decide whether to accept or not
  if(R > runif(1)) {   # if accepted
      mcmc_df$alpha[i] <- cand
      likelihood <- cand_like
    } else {
      mcmc_df$alpha[i] <- mcmc_df$alpha[i-1]
  }
```

Let's go through this line by line. In `step 1a`, we generate a new value of $\alpha$ from a normal distribution centered on the previous value. Remember that the tuning parameter we set earlier determines how big the jumps are between the current and proposed values.

In `step 1b`, we use the `calc_like()` function to estimate the likelihood of our data given the new value of $\alpha$ and the previous values of $\beta$ and $\sigma$. This is what we mean when we say that we treat the other parameters as *fixed* and *known*.

In `step 1c`, we estimate the joint probability of both the current and proposed values by adding the log likelihood to the log prior probability (note that sum of the log values is the same as the product of the probabilities).

In `step 1d`, we estimate the probability of accepting the new value. Note that if the proposed value is more likely than the current value ($R > 1$), this function will return 1. Otherwise, it returns the ratio. The closer the joint probabilities are, the closer $R$ will be to 1. The less likely the proposed value is relative to the current value, the smaller $R$ will be. This means that we will be much less likely to accept new values of $\alpha$ that are a lot less probable than the current value (though not impossible).

Finally, in `step 1e` we accept or reject the proposed value. Note that we compare $R$ to a random value generated from a $Uniform(0,1)$ distribution. If $R=1$, it will always be greater than this value so we will always accept the proposed value (`mcmc_df$alpha[i] <- cand`) and we update the current likelihood to match the current value of $\alpha$. If $R<1$, it turns out testing whether $R$ is less than our randomly generated value from $Uniform(0,1)$ ensures that we accept $\alpha$ with probability $R$ (I'll leave it to you to prove this based on the properties of the uniform distribution).

Next we do the same thing for $\beta$, using the new value of $alpha$ (`mcmc_df$alpha[i]`) and the previous value of $\sigma$ (`mcmc_df$sigma[i - 1]`):

```{r eval = FALSE}
########################
#### 2. Update beta
########################

## 2a: Generate candidate value
  cand <- rnorm(1, mcmc_df$beta[i - 1], tuning)

## 2b: Calculate likelihood at candidate value
  cand_like <- calc_like(y = y, alpha = mcmc_df$alpha[i], beta = cand, 
                         sigma = mcmc_df$sigma[i-1], x = x)
  
## 2c: Calculate likelihood * prior at old value and candidate value
  jointOld <- likelihood + priorBeta(mcmc_df$beta[i-1]) 
  jointCand <- cand_like + priorBeta(cand) 
  
## 2d: Acceptance probability 
  R <- min(1, exp(jointCand - jointOld))

## 2e: Decide whether to accept or not
  if(R > runif(1)) {   # if accepted
      mcmc_df$beta[i] <- cand
      likelihood <- cand_like
    } else {
      mcmc_df$beta[i] <- mcmc_df$beta[i-1]
    }
```

Finally, we update $\sigma$ using the new values of both $\alpha$ and $\beta$. The only difference here is that because $\sigma$ has to be $>0$, the normal proposal distribution we used for $\alpha$ and $\beta$ will not work (it could create negative values). As we learned in lecture, you could create a proposal distribution that generates positive value (e.g., gamma) and use moment matching to estimate the parameters with regards to the current $\sigma$ and the tuning parameter. We'll use a slightly less efficient approach here to demonstrate there are different ways to enforcing the behavior we want in the sampler. In this cases, will simply add a small value to the current $\sigma$ and automatically reject the proposed value if it's less than 0. The small value can be negative or positive to allow us to explore the posterior.

```{r eval = FALSE}
########################
#### 3. Update sigma
########################

## 3a: Generate candidate value
    cand <- mcmc_df$sigma[i-1] + runif(1,  -.1, 0.1)
    # If candidate value is outside [0,Inf], keep the old value of sigma
    if (cand < 0) {
      mcmc_df$sigma[i] <- mcmc_df$sigma[i-1]
    } else {
## 3b: Calculate likelihood at candidate value
      cand_like <- calc_like(y = y, alpha = mcmc_df$alpha[i], beta = mcmc_df$beta[i], 
                          sigma = cand, x = x)
      
## 3c: Calculate likelihood * prior at old value and candidate value
      jointOld <- likelihood + priorsigma(mcmc_df$sigma[i-1])
      jointCand <- cand_like + priorsigma(cand)

## 3d: Acceptance probability 
      R <- min(1, exp(jointCand - jointOld))

## 3e: Decide whether to accept or not
      if(R > runif(1)) {   # if accepted
        mcmc_df$sigma[i] <- cand
        likelihood <- cand_like
      } else {
        mcmc_df$sigma[i] <- mcmc_df$sigma[i-1]
      }
    }
  }

```

# Running the sampler

Finally, we can run the sampler. Create a new script called `metroplis.R` (or something similar) and put it in `scripts/`. First, we will load the packages we need and use the `source()` function to read in our custom functions (`source()` runs the `.R` files given in the argument which in this case will make the functions available for our sampler):

```{r eval = FALSE}
library(WILD6900)

source("R/calc_like.R")
source("R/priors.R")
```

Next, read in the simulated data and set the initial values for the MCMC:

```{r eval = FALSE}
## Read simulated data
  sim_dat <- readRDS("data/sim_seed_counts.rds")

## Length of chains
  nIter <- 10000

## Tuning parameter
  tuning <- 1.5

## Empty data frame to store posterior samples of each parameter
  mcmc_df <- data.frame(x = 1:nIter,
                        alpha = numeric(nIter),
                        beta = numeric(nIter),
                        sigma = numeric(nIter),
                        accept.alpha = 0,
                        accept.beta = 0, 
                        accept.sigma = 0)

## Initial values
  mcmc_df$alpha[1] <-  runif(1, 200, 300)
  mcmc_df$beta[1] <- runif(1, 25, 75)
  mcmc_df$sigma[1] <- runif(1, 0, 10)
  
## Initial likelihood
  likelihood <- calc_like(y = sim_dat$y, alpha = mcmc_df$alpha[1], 
                       beta = mcmc_df$beta[1], sigma = mcmc_df$sigma[1], 
                       x = sim_dat$visits.c) 
```

```{r echo = FALSE}
data("sim_seed_counts")
sim_dat <- sim_seed_counts

## Length of chains
  nIter <- 10000

## Tuning parameter
  tuning <- 1.5

## Empty data frame to store posterior samples of each parameter
  mcmc_df <- data.frame(x = 1:nIter,
                        alpha = numeric(nIter),
                        beta = numeric(nIter),
                        sigma = numeric(nIter),
                        accept.alpha = 0,
                        accept.beta = 0, 
                        accept.sigma = 0)

## Initial values
  mcmc_df$alpha[1] <-  runif(1, 200, 300)
  mcmc_df$beta[1] <- runif(1, 25, 75)
  mcmc_df$sigma[1] <- runif(1, 0, 10)
  
## Initial likelihood
  likelihood <- calc_like(y = sim_dat$y, alpha = mcmc_df$alpha[1], 
                       beta = mcmc_df$beta[1], sigma = mcmc_df$sigma[1], 
                       x = sim_dat$visits.c) 
```

Now take the sampler code and put it inside a `for` loop to create the chains. Note that in the likelihood functions we will need to reference our simulated data and covariates:

```{r}
for(i in 2:nIter){
########################
#### 1. Update alpha
########################

## 1a: Generate candidate value
  cand <- rnorm(1, mcmc_df$alpha[i - 1], tuning)

## 1b: Calculate likelihood at candidate value
  cand_like <- calc_like(y = sim_dat$y, alpha = cand, 
                         beta = mcmc_df$beta[i-1], sigma = mcmc_df$sigma[i-1], 
                         x = sim_dat$visits.c)
  
## 1c: Calculate likelihood * prior at old value and candidate value
  jointOld <- likelihood + priorAlpha(mcmc_df$alpha[i-1]) 
  jointCand <- cand_like + priorAlpha(cand) 
  
## 1d: Acceptance probability 
  R <- min(1, exp(jointCand - jointOld))

## 1e: Decide whether to accept or not
  if(R > runif(1)) {   # if accepted
      mcmc_df$alpha[i] <- cand
      likelihood <- cand_like
      mcmc_df$accept.alpha[i] <- 1
    } else {
      mcmc_df$alpha[i] <- mcmc_df$alpha[i-1]
    }
  
########################
#### 2. Update beta
########################

## 2a: Generate candidate value
  cand <- rnorm(1, mcmc_df$beta[i - 1], tuning)

## 2b: Calculate likelihood at candidate value
  cand_like <- calc_like(y = sim_dat$y, alpha = mcmc_df$alpha[i], beta = cand, 
                         sigma = mcmc_df$sigma[i-1], x = sim_dat$visits.c)
  
## 2c: Calculate likelihood * prior at old value and candidate value
  jointOld <- likelihood + priorBeta(mcmc_df$beta[i-1]) 
  jointCand <- cand_like + priorBeta(cand) 
  
## 2d: Acceptance probability 
  R <- min(1, exp(jointCand - jointOld))

## 2e: Decide whether to accept or not
  if(R > runif(1)) {   # if accepted
      mcmc_df$beta[i] <- cand
      likelihood <- cand_like
      mcmc_df$accept.beta[i] <- 1
    } else {
      mcmc_df$beta[i] <- mcmc_df$beta[i-1]
    }
  
########################
#### 3. Update sigma
########################

## 3a: Generate candidate value
    cand <- mcmc_df$sigma[i-1] + runif(1,  -1, 1)
    # If candidate value is outside [0,Inf], keep the old value of sigma
    if (cand < 0) {
      mcmc_df$sigma[i] <- mcmc_df$sigma[i-1]
    } else {
## 3b: Calculate likelihood at candidate value
      cand_like <- calc_like(y = sim_dat$y, alpha = mcmc_df$alpha[i], beta = mcmc_df$beta[i], 
                          sigma = cand, x = sim_dat$visits.c)
      
## 3c: Calculate likelihood * prior at old value and candidate value
      jointOld <- likelihood + priorsigma(mcmc_df$sigma[i-1])
      jointCand <- cand_like + priorsigma(cand)

## 3d: Acceptance probability 
      R <- min(1, exp(jointCand - jointOld))

## 3e: Decide whether to accept or not
      if(R > runif(1)) {   # if accepted
        mcmc_df$sigma[i] <- cand
        likelihood <- cand_like
        mcmc_df$accept.sigma[i] <- 1
      } else {
        mcmc_df$sigma[i] <- mcmc_df$sigma[i-1]
      }
    }
}
```

At this point, you should be able to run the script from start to finish and create the posterior samples for each parameter.

# Checking the output

Now we can determine whether the sampler returns the data generating values from our data simulation. When assessing the output of MCMC chains, a good first diagnostic is usually checking the trace plots:

```{r fig.width=7, fig.height=4}
ggplot(mcmc_df, aes(x = x, y = alpha)) + geom_path()
```

The chain for $\alpha$ moved rapidly from the initial value to the stationary posterior distribution but those first few samples will obviously bias our parameter estimates if we include them in the posterior. Let's remove them by using the `dplyr` function `slice()`:

```{r fig.width=7, fig.height=4}
mcmc_df <- dplyr::slice(mcmc_df, 1000:nIter)

ggplot(mcmc_df, aes(x = x, y = alpha)) + geom_path()
```

That looks better. Now we can estimate the mean and 95% credible interval of the posterior:

```{r}
mean(mcmc_df$alpha)
quantile(mcmc_df$alpha, probs = c(0.025, 0.975))
```

Pretty close to the data generating value of 250. Now let's check $\beta$:

```{r fig.width=7, fig.height=4}
ggplot(mcmc_df, aes(x = x, y = beta)) + geom_path()
```

and

```{r}
mean(mcmc_df$beta)
quantile(mcmc_df$beta, probs = c(0.025, 0.975))
```

Also pretty close to the data generating value of 50. Finally, $\sigma$

```{r fig.width=7, fig.height=4}
ggplot(mcmc_df, aes(x = x, y = sigma)) + geom_path()

mean(mcmc_df$sigma)
quantile(mcmc_df$sigma, probs = c(0.025, 0.975))
```

Wasn't that easy?

# Homework

1.  In simulation, we set $\alpha$ = 250, and we provided a reasonable, diffuse prior in the example above. But what happens if we give a restrictive (and wrong) prior for $\alpha$? Change your `priorAlpha` function to instead match $\alpha \sim Uniform(0, 150)$. Make sure to also change this line to match the new prior:

```{r, echo = T, eval = F}
mcmc_df$alpha[1] <-  runif(1, 200, 300) #change this to something reasonable
```

Run your sampler again and print out the chain for $\alpha$. What do you see?

2.  Go back to the original prior for alpha. We didn't discuss it much in lab, but the tuning parameter affects how fast the sampler will explore the state space. If it's really small, it might take forever. If it's really large, it can make giant jumps and reject a lot of possible values, which wastes time and causes the sampler to 'stall'. We can see what the acceptance rate of our sampler was by calculating the mean of the accept.alpha, accept.beta and accept.sigma columns. Run the sampler with a tuning parameter of .05 and 50. Calculate the acceptance rates for each of the 3 parameters. What do you notice?

# Homework (+1 Bonus Point)

3.  When we generated the seed count data, we did so assuming a linear model that linked pollination visits to seed counts: $y_i = \alpha + \beta * x_i + \epsilon_i$. Because we used simulated data, we know this model ahead of time. However in real life, we wouldn't know the true model ahead of time and might try to fit a different model to the data. Adding on to what we already have, create an MCMC that also includes a second beta term that looks at visits squared: $y_i = \alpha + \beta * x_i + \beta_2 * x_i^2 + \epsilon_i$. Run the MCMC and look at the output. What does the MCMC estimate for $\beta_2$?

> Hint: Approach this problem one step at a time.
>
> \- Begin by changing the mcmc_df to allow you to store an additional beta parameter.
>
> \- Next, choose an initial value for beta2.
>
> \- Change your calc_like function to include beta2. Remember that lp will just be alpha + beta*x + beta2*x\*x.
>
> \- Optionally, create another function for priorBeta2 (or just re-use the priorBeta function for beta2 as well).
>
> Next create a section in your sampler to update Beta2. It will look almost identical to the one already present for Beta.
>
> \- Run your model.
>
> \- Finally, use the quantile function to look at the estimate: quantile(mcmc_df\$beta2, probs = c(0.025,0.5, 0.975))

4.  On a 1-10 scale, with 1 being the worst week ever and 10 being the best, how would you rate this week's content? What lingering questions/confusion about the lecture or lab do you still have?
