---
title: "Lab9: Spatial Capture Recapture"
output: 
  rmarkdown::html_vignette:
    keep_md: yes
vignette: >
  %\VignetteIndexEntry{Lab8_Movement}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
   warning = FALSE, message = FALSE,
   fig.align = 'center', fig.retina = 2,
  collapse = TRUE,
  comment = "#>"
)
library(dplyr)
library(knitr)
source(here::here("R/zzz.R"))
```

```{r setup, echo=FALSE}
library(WILD8370)
library(nimble)
library(coda)
library(mcmcOutput)
library(MCMCvis)
library(terra)
```

In this activity we'll practice running both a spatial and non-spatial mark recapture model on some data. 


## Butterflies!

Today's dataset is on a study of false heath fritillary butterflies (*Melitaea diamina*) conducted in Finland. You can find more details about this study here:

Fabritius, H., Rönkä, K. & Ovaskainen, O. The dual role of rivers in facilitating or hindering movements of the false heath fritillary butterfly. Mov Ecol 3, 4 (2015). https://doi.org/10.1186/s40462-015-0031-z

The full study uses three datasets but today we will be only using the data from 1995. 

From the authors:

Details of the three data sets summarized below. The number of search days without (with) parenthesis indicates the length of study period (versus the number of effective search days with search activity).

+ Location: Siitama, Pirkanmaa, Finland (61.6°N, 24.2°E)
+ Year of data collection: 1995
+ Coordinator: Niklas Wahlberg
+ Coordinate reference system: Finland Zone 3 (EPSG:2393)

We will work with 3 files to get the information we need:

1. Effort log - a 1/0 matrix indicating if the 'trap' was operational 
2. Records of each time an individual was captured (and what trap)
3. Trap locations in Finland Zone 3 units 

## Non-Spatial CMR

Let's begin by analyzing the data in a non-spatial framework. We know that sometimes not all areas were surveyed, so we'll model detection probability as a function of the number of traps that were open.

```{r}
nimbleCode({
  psi ~ dbeta(1, 1) #N/M
  for(t in 1:nocc){
    logit(p[t]) = alpha0 + alpha1*effort[t]
  }
  
  for(i in 1:M){
    z[i] ~ dbern(psi)
    
    for(t in 1:nocc){
      y[i, t] ~ dbern(p[t] * z[i])
    } # end t
  } # end i
  
  N <- sum(z[1:M])
})
```


Let's grab the data to get it into the format we need:

```{r}
effort <- Finnish_butterflies$effort
head(effort)
```
There appear to be 14 trap occasions and 112 traps. 

```{r}
data('butterflies')
caps <- as.data.frame(Finnish_butterflies$cap_hist)
head(caps)
```
The data gives us one row per capture. Now we need to put this into one row per individual, one column per occasion. 

```{r}
nind <- length(unique(caps$Individual))
nocc <- 14
y <- array(0, c(nind, nocc))
for(i in 1:nrow(caps)){
  y[caps$Individual[i],caps$Day[i]] <- 1
}
head(y)
```




## Homework

1. 

2. 

3.  On a 1-10 scale, with 1 being the worst week ever and 10 being the best, how would you rate this week's content? What lingering questions/confusion about the lecture or lab do you still have?
