---
title: "Lecture 9b"
subtitle: "Static Occupancy Models"
author: "<br/><br/><br/>Spring 2025"
output:
  xaringan::moon_reader:
    css: ["default", "FANR8370.css", "FANR8370-fonts.css"]
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
      ratio: '16:9'
      slideNumberFormat: '%current%' 
---

```{r setup, include = FALSE}
options(htmltools.dir.version = FALSE)
knitr::opts_chunk$set(echo = FALSE, fig.align = 'center', warning=FALSE, message=FALSE,fig.retina = 2)
library(WILD8370)
library(nimble)
library(MCMCvis)
source(here::here("R/zzz.R"))
# library(gganimate)
```

# Readings: 
<br/>
> Kéry & Schaub 383-411

<br/>

> Occupancy Estimation and Modeling : Inferring Patterns and Dynamics of Species Occurrence by Darryl I. MacKenzie, James D. Nichols, J. Andrew Royle, Kenneth H. Pollock, Larissa Bailey, and James E. Hines


---
## Estimating abundance? 

#### For the past few weeks, we've been modeling abundance:

$$\Large N_t \sim Poisson(\lambda)$$

--
#### Occupancy is the probability a site is occupied

- Occupancy comes free from abundance
    + If $\large N_i > 0$, $\large z_i = 1$

    + If $\large N_i = 0$, $\large z_i = 0$

--
#### So even when occupancy is the state-variable, we are still modeling something related to abundance

---
## Estimating abundance? 

#### Typical state model for occupancy

$$\Large z_i \sim Bernoulli(\psi)$$

--
#### If the expected abundance is $\large \lambda$, what is the probability $\large N = 0$?

$$\Large Pr(N=0|\lambda)=\frac{\lambda^0e^-\lambda}{0!}=e^{-\lambda}$$ 

--
#### If the expected abundance is $\large \lambda$, what is the probability $\large N > 0$?

$$\Large 1 - Pr(N=0|\lambda)=1 -e^{-\lambda}$$ 

--
#### So (if our assumptions are valid):

$$\Large \psi = 1 - e^{-\lambda}$$

---
## Why estimate occupancy?

#### Less effort

--
#### Historical data sets

--
#### More reliable when $\large N$ very small

--
#### Occupancy = abundance (e.g., breeding territory)

--
#### Metapopulation dynamics

--
#### Distribution/range size

--
#### Disease dynamics

---
## Why not just use observed presence/absence?

#### As in all ecological studies, we rarely (if ever) observe the state process perfectly 

--
#### In the case of occupancy, some sites will be occupied but we will fail to detect the species

- i.e., $\large p < 1$


--
#### Also possible (though generally more rare) that we record the species when it's not present (false positive)

- see Royle & Link 2006


--
#### Similar to N-mixture models, estimating $\large p$ requires temporal replication


---
## Single-season (static) occupancy model

#### State-space formulation

- State-model

$$\Large z_i \sim Bernoulli(\psi_i)$$

$$\Large logit(\psi_i) = \alpha_0 + \alpha_1x_i$$

- Observation-model

$$\Large y_{ik} \sim Bernoulli(z_i \times p_{ik})$$


$$\Large logit(p_{ik}) = \beta_0 + \beta_1x_{ik}$$

---
## Assumptions of closed occupancy models
<br/>

1) Binomial distribution is a true description of state/observation processes

<br/>

2) Occupancy at each site is random and independent of occupancy at all other sites

<br/>

3) Population is closed between surveys

<br/>

4) Observers do not produce false positives

---
## Estimating $\LARGE p$

#### Imagine a single site surveyed 3 times:

- Assume site is closed across samples

- Assume constant $\large p$

$$\LARGE y_i = [111]$$

--
#### What is the likelihood of this observation?

--
P(occupied) x P(observed 3 times)

<br/>

$$\LARGE \psi p^3$$



---
## Estimating $\LARGE p$

#### What about?

$$\LARGE y_i = [011]$$

--
P(occupied) x P(not observed 1 time) x P(observed 2 times)

<br/>
$$\LARGE \psi (1-p)p^2$$
---
## Estimating $\LARGE p$

#### Is this scenario any different?

$$\LARGE y_i = [101]$$

--
No, the order of detection/non-detection does not matter.

<br/>
P(occupied) x P(not observed 1 time) x P(observed 2 times)

<br/>
$$\LARGE \psi (1-p)p^2$$


---
## Estimating $\LARGE p$

#### What about?

$$\LARGE y_i = [000]$$

--
<br/>
Three 0's can arise from two different scenarios - the site is not occupied, or the site is occupied but we missed it 3 times. 

--

<br/>
P(not occupied) + P(occupied) x P(not observed 3 times)

<br/>

$$\LARGE (1 - \psi) + \psi (1-p)^3$$

---

## In Class Exercise 

Simulate 15 years of an occupancy system using the following information. Occupancy can change between years, but this change is not modeled:
 
- `set.seed(55)` 

- $\psi$ = .32

- nsites = 40

- Binomial detection process with 3 site visits per year. 

-  $E(p)$ = .4 + $\eta_t$ 

-  $\eta_t \sim Normal(0, 1)$

---

## In Class Exercise 

```{r, echo = T}
set.seed(55)
nyears <- 15
psi <- .32
nsites <- 40
nvisits <- 3
mean_p <- .4
p <- plogis(mean_p + rnorm(nyears, 0, 1))

z <- y <-array(NA, c(nsites, nyears))
for(t in 1:nyears){
  for(i in 1:nsites){
    z[i,t] <- rbinom(1, 1, psi)
    y[i,t] <- rbinom(1, z[i,t]*nvisits, p[t])
  } #end i
} #end t

```

---

## In Class Exercise 


What proportion of sites were occupied in each year?

--

```{r, eval = T, echo = T}
colSums(z)/nsites
```

<br/>

--

How about observed occupancy? 

--

```{r, echo = T}
colSums(y > 0)/nsites
```

---
# Expected vs Realized Occupancy

```{r, fig.height = 5.5}
gg_1 <- data.frame(Year = 1:nyears,
                   Occupancy = c(rep(psi, nyears), colSums(z)/nsites, colSums(y > 0)/nsites),
                   Type = rep(c('Expected', 'Realized', 'Observed'), each = nyears))
ggplot(gg_1, aes(x = Year, y = Occupancy, col = Type))+
  geom_line()+
  geom_point(cex = 5)+
  theme_classic()+
  theme(axis.text = element_text(size = 15), axis.title = element_text(size = 20), legend.text = element_text(size = 15), legend.title = element_text(size = 20))
```

---
## Nimble Model

```{r, eval = T, echo =T}
occ.mod <- nimbleCode({
for(t in 1:nyears){
  for(i in 1:nsites){
    z[i,t] ~dbern(psi)
    y[i,t] ~dbinom(p = p[t], size = z[i,t]*nvisits)
  } #end i 
  
  logit(p[t]) <- p0+p1[t]
  p1[t] ~ dnorm(0, 1)
} #end t
  
psi ~ dbeta(1, 1)
p0 ~ dnorm(0, 1)
})
```

Quiz- is `p1` being treated as a random or fixed effect? 

---
## Nimble Model

```{r, eval = T, echo =T}
params <- c('p', 'p0', 'p1', 'psi')
n.c <- list(nsites = nsites, nvisits = nvisits, nyears = nyears)
n.d <- list(y =y)
n.i <- function()list(p0 = rnorm(1), p1 = rnorm(nyears), psi = runif(1))
```

--

Initial values for $z$ can be tricky - we need to make sure we don't suggest a site is unoccupied if we had observations there 
<br/>
--
One option is to start anywhere with y = 0 as unoccupied

```{r, eval = T, echo =T}
z.init <- (y >0)*1
n.i <- function()list(p0 = rnorm(1), p1 = rnorm(nyears), psi = runif(1), z =z.init)
```

--
Alternatively, flip a coin for any sites where y = 0
```{r, eval = T, echo =T}
z.init <- matrix(rbinom(nsites*nyears, 1, .5), ncol = nyears)
z.init[which(y > 0)] <- 1 
n.i <- function()list(p0 = rnorm(1), p1 = rnorm(nyears), psi = runif(1), z =z.init)
```

---
## Run the model in Nimble

```{r, echo = T, eval = F}
occ_out <- nimbleMCMC(code = occ.mod,
                     data = n.d,
                     constants = n.c,
                     inits = n.i(),
                     monitors = params,
                     thin = 1,
                     niter = 5000,
                     nburnin = 2500,
                     nchains = 3,
                     samplesAsCodaMCMC = TRUE
                      )
```

```{r}
occ_out <- readRDS('occ_out1.rds')
```

---
## Check the model
```{r, eval=T, fig.height=4.5, fig.width=8}
MCMCvis::MCMCtrace(occ_out, params = c('p0', 'psi'), pdf = F)
```

---
## Compare with simulation
```{r, eval=T, fig.height=5.5, fig.width=8}
est <- (summary(occ_out)$quantiles[1:15,])
est_gg <- data.frame(median = est[,3],
                     LCI = est[,1],
                     UCI = est[,5],
                     time = 1:nyears)
true_gg <- data.frame(p = p, time = 1:nyears)
ggplot()+
  geom_pointrange(data =est_gg, aes(x = time, y= median, ymin = LCI, ymax = UCI))+
  geom_point(data= true_gg, aes(x = time, y= p), pch = 8, col = 'red')+
  theme_bw()+
  xlab('Time')+
  ylab('Detection Probability')+
  theme(axis.text= element_text(size =15), axis.title = element_text(size =20))
  
```
