---
title: "Lecture 10"
subtitle: "Dynamic Occupancy Models"
author: "<br/><br/><br/>Spring 2025"
output:
  xaringan::moon_reader:
    css: ["default", "FANR8370.css", "FANR8370-fonts.css"]
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
      ratio: '16:9'
      slideNumberFormat: '%current%' 
---

```{r setup, include = FALSE}
options(htmltools.dir.version = FALSE)
knitr::opts_chunk$set(echo = FALSE, fig.align = 'center', warning=FALSE, message=FALSE,fig.retina = 2)
library(WILD8370)
library(nimble)
library(MCMCvis)
source(here::here("R/zzz.R"))
# library(gganimate)
```

# Readings: 
<br/>
> Kéry & Schaub 383-411



---
## Closed Occupancy models
<br/>

$$\Large logit(\psi_j) = \beta_0 + \beta_1*x_j $$ 

<br/>

$$ \Large z_j \sim Bernoulli(\psi_j) $$ 

<br/>


$$ \Large y_{j,k} \sim Bernoulli(z_j, p_k)$$
---

## Assumptions of closed occupancy models
<br/>

1) Binomial distribution is a true description of state/observation processes

<br/>

2) Occupancy at each site is random and independent of occupancy at all other sites

<br/>

3) Population is closed between surveys

<br/>

4) Observers do not produce false positives

---
## Estimating $\LARGE p$

#### Imagine a single site surveyed 3 times:

- Assume site is closed across samples

- Assume constant $\large p$

$$\LARGE y_i = [111]$$

--
#### What is the likelihood of this observation?

--
P(occupied) x P(observed 3 times)

<br/>

$$\LARGE \psi p^3$$



---
## Estimating $\LARGE p$

#### What about?

$$\LARGE y_i = [011]$$

--
P(occupied) x P(not observed 1 time) x P(observed 2 times)

<br/>
$$\LARGE \psi (1-p)p^2$$


---
## Estimating $\LARGE p$

#### What about?

$$\LARGE y_i = [000]$$

--
<br/>
Three 0's can arise from two different scenarios - the site is not occupied, or the site is occupied but we missed it 3 times. 

--

<br/>
P(not occupied) + P(occupied) x P(not observed 3 times)

<br/>

$$\LARGE (1 - \psi) + \psi (1-p)^3$$

---
## Multi-season (dynamic) occupancy model

#### What if occupancy can change over time? 

- Data collection using the *robust design*
    + Population open between primary periods (e.g., years)
    + Population closed within secondary periods (e.g., occasions)

<br/>

$$\Large y_i = [\underbrace{101}_{Year\;1}\;\;\;\;\; \underbrace{000}_{Year\;2}\;\;\;\;\;\underbrace{110}_{Year\;3}\;\;\;\;\;\underbrace{100}_{Year\;4}]$$

<br/>
- We need a model to explain *why* occupancy can change

---
## Occupancy Dynamics

#### What causes changes in occupancy?

Assume some site is occupied in year 1:
```{r, warning = F, message=FALSE}
library(ggplot2)

g <- data.frame(x = c(0,1, -1),
                y = c(1,1,1))

g <- graph_from_data_frame(
  d = data.frame(from = c(1, 1), to = c(2, 3), label = c('1-\U03C6', '\U03C6')), 
  directed = TRUE,
  vertices = data.frame(name = c(1, 2, 3), label = c('1', '0', '1'))
)

```

```{r, eval = F, echo = F}
library(DiagrammeR)
node_list <- data.frame(id = 1:3, label = c(1,0,1))
edge_list <- data.frame(from = c(1,1), to = c(2,3), label = c('1-\\U03C6', '<$\\phi$>'))
igraph1 <- create_graph() %>% 
  add_nodes_from_table(
    table = node_list,
    label_col = label
  ) %>%
  add_edges_from_table(
    table = edge_list,
    from_col = from,
    to_col = to,
    from_to_map = id_external
  ) %>%
  select_nodes(conditions = label == 1) %>%
  set_node_attrs_ws(node_attr = fillcolor, value = "orange") %>%
  clear_selection()
render_graph(igraph1, layout = 'tree', node_attrs = 'labelloc = t')
```
```{r, eval = T, echo = F, fig.width= 3.5}

a_graph <-
   create_graph() %>%
   add_node(node_aes = node_aes(
       color = "steelblue",
       fillcolor = "lightblue",
       fontcolor = "gray35"
     ))
render_graph(a_graph)
```
---
#### What causes changes in occupancy?

In year 2, it can either stay occupied (survive) or no longer be occupied:

```{r, eval = T, echo = F, fig.width= 3.5}
library(DiagrammeR)
a_graph <-
   create_graph() %>%
   add_node(node_aes = node_aes(
       color = "steelblue",
       fillcolor = "lightblue",
       fontcolor = "gray35"
     ))%>%
  add_node() %>%
  add_edge(from = 1, to = 2)
render_graph(a_graph)
```
---

---
## Multi-season (dynamic) occupancy model

- In year 1:
$$\Large z_{i,1} \sim Bernoulli(\psi)$$

- In years 2+:
$$\Large z_{i,t} \sim Bernoulli(z_{i,t-1}(1-\epsilon)+(1-z_{i,t-1}\gamma))$$

- $\Large z_{i,t}$ Binary occupancy state
- 


