---
title: "Lecture 10"
subtitle: "Dynamic Occupancy Models"
author: "<br/><br/><br/>Spring 2025"
output:
  xaringan::moon_reader:
    css: ["default", "FANR8370.css", "FANR8370-fonts.css"]
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
      ratio: '16:9'
      slideNumberFormat: '%current%' 
---

```{r setup, include = FALSE}
options(htmltools.dir.version = FALSE)
knitr::opts_chunk$set(echo = FALSE, fig.align = 'center', warning=FALSE, message=FALSE,fig.retina = 2)
library(WILD8370)
library(nimble)
library(MCMCvis)
source(here::here("R/zzz.R"))
# library(gganimate)
```

# Readings: 
<br/>
> Kéry & Schaub 383-411

<br/>

> Occupancy Estimation and Modeling : Inferring Patterns and Dynamics of Species Occurrence by Darryl I. MacKenzie, James D. Nichols, J. Andrew Royle, Kenneth H. Pollock, Larissa Bailey, and James E. Hines



---
## Closed Occupancy models
<br/>

$$\Large logit(\psi_j) = \beta_0 + \beta_1*x_j $$ 

<br/>

$$ \Large z_j \sim Bernoulli(\psi_j) $$ 

<br/>


$$ \Large y_{j,k} \sim Bernoulli(z_j, p_k)$$
---

## Assumptions of closed occupancy models
<br/>

1) Binomial distribution is a true description of state/observation processes

<br/>

2) Occupancy at each site is random and independent of occupancy at all other sites

<br/>

3) Population is closed between surveys

<br/>

4) Observers do not produce false positives

---
## Estimating $\LARGE p$

#### Imagine a single site surveyed 3 times:

- Assume site is closed across samples

- Assume constant $\large p$

$$\LARGE y_i = [111]$$

--
#### What is the likelihood of this observation?

--
P(occupied) x P(observed 3 times)

<br/>

$$\LARGE \psi p^3$$



---
## Estimating $\LARGE p$

#### What about?

$$\LARGE y_i = [011]$$

--
P(occupied) x P(not observed 1 time) x P(observed 2 times)

<br/>
$$\LARGE \psi (1-p)p^2$$


---
## Estimating $\LARGE p$

#### What about?

$$\LARGE y_i = [000]$$

--
<br/>
Three 0's can arise from two different scenarios - the site is not occupied, or the site is occupied but we missed it 3 times. 

--

<br/>
P(not occupied) + P(occupied) x P(not observed 3 times)

<br/>

$$\LARGE (1 - \psi) + \psi (1-p)^3$$

---
## Multi-season (dynamic) occupancy model

#### What if occupancy can change over time? 

- Data collection using the *robust design*
    + Population open between primary periods (e.g., years)
    + Population closed within secondary periods (e.g., occasions)

<br/>

$$\Large y_i = [\underbrace{101}_{Year\;1}\;\;\;\;\; \underbrace{000}_{Year\;2}\;\;\;\;\;\underbrace{110}_{Year\;3}\;\;\;\;\;\underbrace{100}_{Year\;4}]$$

---
## Probability of Observations

#### What is the likelihood of this observation?

$$\Large y_i = [\underbrace{101}_{Year\;1}\;\;\;\;\; \underbrace{000}_{Year\;2}\;\;\;\;\;\underbrace{110}_{Year\;3}]$$
--

Year 1:  $$\psi_1p_1^2(1-p_1)$$
<br/>
--
Year 2:  $$\psi_2(1-p_2)^3 + (1-\psi_2)$$
<br/>
--
Year 3:  $$\psi_3p_3^2(1-p_3)$$

---
## Probability of Observations

#### What is the likelihood of this observation?

$$\Large y_i = [\underbrace{101}_{Year\;1}\;\;\;\;\; \underbrace{000}_{Year\;2}\;\;\;\;\;\underbrace{110}_{Year\;3}]$$
Therefore $$Pr(y_i = [101\;\;\;\;\; 000 \;\;\;\;\;110 \;\;\;\;\; 100 ]| \boldsymbol{ \psi, p} ) = \\\psi_1p_1^2(1-p_1)\\\times\psi_2(1-p_2)^3 + (1-\psi_2)\\\times\psi_3p_3^2(1-p_3)$$


---
## Occupancy Dynamics

Remember that occupancy is a *markovian* process - therefore occupancy status is not independent between years! We need a model to explain *why* occupancy can change.

Assume some site is occupied in year 1. In year 2, it can either stay occupied (survive) or no longer be occupied (go 'extinct')

```{r, warning = F, message=FALSE}
library(ggplot2)

g <- data.frame(x = c(0,1, -1),
                y = c(3,1,1),
                occ = c('1', '1', '0'))
lines <- data.frame(x = c(0, -.825, 0, .825),
                    y = c(3,1.35,3,1.35))

a <- ggplot(g, aes(x = x, y= y))+
  geom_line(data = lines[1:2,], arrow = arrow(angle = 15, type = 'closed', ends = 'both'))+
  geom_line(data = lines[3:4,], arrow = arrow(angle = 15, type = 'closed', ends = 'both'))+
  geom_point(cex = 30, aes(col = occ))+
  geom_text(aes(label = occ), size = 15)+
  geom_label(x = -.5, y = 2, label = "1-phi", parse = TRUE, size = 8)+
   geom_label(x = .5, y = 2, label = "phi", parse = TRUE, size =8)+
  theme_void()+
  xlim(-2, 2)+
  ylim(0,4)+
  theme(legend.position = 'none')

g2 <- data.frame(x = c(0,1, -1),
                y = c(3,1,1),
                occ = c('0', '1', '0'))

b <- ggplot(g2, aes(x = x, y= y))+
  geom_line(data = lines[1:2,], arrow = arrow(angle = 15, type = 'closed', ends = 'both'))+
  geom_line(data = lines[3:4,], arrow = arrow(angle = 15, type = 'closed', ends = 'both'))+
  geom_point(cex = 30, aes(col = occ))+
  geom_text(aes(label = occ), size = 15)+
  geom_label(x = -.5, y = 2, label = "1-gamma", parse = TRUE, size = 8)+
   geom_label(x = .5, y = 2, label = "gamma", parse = TRUE, size = 8)+
  theme_void()+
  xlim(-2, 2)+
  ylim(0,4)+
  theme(legend.position = 'none')
```

```{r, echo = F, eval=T, include= T, fig.height = 6.25}
a
```

---
## Occupancy Dynamics

Similarly, a site can be empty in year 1. In year 2, it can either become occupied (colonization) or remain unoccupied. 

```{r, echo = F, eval=T, include= T, fig.height = 6.25}
b
```

---
## Multi-season (dynamic) occupancy model

- In year 1:
$$\Large z_{i,1} \sim Bernoulli(\psi)$$
<br/>

- In years 2+:
$$\Large z_{i,t} \sim Bernoulli(z_{i,t-1}(1-\phi)+(1-z_{i,t-1})\gamma)$$


- $\Large z_{i,t}$ = Binary occupancy state

- $\Large\psi$ = Initial occupancy probability 

- $\Large\phi$ = Survival probability

- $\Large\gamma$ = Colonization probability 

*Some authors will use different greek letters 

---
## Multi-season (dynamic) occupancy model

We can easily model any of our probabilities ($\psi$,  $\phi$,  $\gamma$) as functions of covariates. 

$$logit(\Large\psi_i) = \beta_0 + \beta_1\boldsymbol{x_i}$$
$$logit(\Large\phi_t) = \alpha_0 + \alpha_1\boldsymbol{x_t}  ...  $$

---
# Probability of Occurrence

There are several derived quantites that are often of interest. 

<br/> 

Probability of occurrence is just the mean $\psi$ at average covariate conditions. We calculate this via $\phi$ and $\gamma$. 

$$\Large \hat{\psi_1} = logit(\beta_0)$$
For $t$ > 1

$$\Large \hat{\psi_t} = \hat{\psi}_{t-1}\phi_{t-1} + (1-\hat{\psi}_{t-1})\gamma_{t-1}$$
---
# Turnover 

Turnover is another common derived quantity. This can be defined multiple ways, though the most common is the proportion of all sites that *just* became occupied relative to the total number of occupied sites. 

<br/> 

You can calculate this as either expected or realized turnover, though expected is more common.  

<br/>

$$\Large \tau_{t} = \frac{(1-\hat{\psi}_{t})\gamma_{t}}{\hat{\psi}_{t}} =\frac{(1-\hat{\psi}_{t})\gamma_{t}}{\hat{\psi}_{t}\phi_{t} + (1-\hat{\psi}_{t})\gamma_{t}}$$

---
# Growth Rate

Finally, some authors like to include a growth rate parameter, which is just the rate of change in occupancy probability over time. 

<br/>

$$\Large \lambda_t = \frac{\hat{\psi_t}}{\hat{\psi}_{t-1}}$$

---


## In Class Exercise 

Simulate 15 years of a dynamic occupancy system using the following information:

- `set.seed(55)` 

- $\psi_1$ = .62

- $\phi$ = .8 

- $\gamma$ = .2

- nsites = 40

- Binomial detection process with $p$ = .4 and 4 site visits per year

---

## In Class Exercise 

```{r, echo = T}
set.seed(55)
nyears <- 15
psi1 <- .62
phi <- .8
gamma <- .1
nsites <- 40
nvisits <- 4
p <- .4

z <- y <-array(NA, c(nsites, nyears))
z[,1] <- rbinom(nsites, 1, psi1)
for(i in 1:nsites){
  y[i,1] <- rbinom(1, z[i,1]*nvisits, p)
    for(t in 2:nyears){
      z[i, t] <- rbinom(1, 1, z[i,t-1]*(1-phi)+(1-z[i,t-1])*gamma)
      y[i, t] <- rbinom(1, z[i,t]*nvisits, p)
  }
}

```

---

## In Class Exercise 


What proportion of sites were occupied in each year?

--

```{r, eval = T, echo = T}
colSums(z)/nsites
```

<br/>

--

How about observed occupancy? 

--

```{r, echo = T}
colSums(y > 0)/nsites
```

---
## In Class Exercise 

What's the expected probability of occurence? 

$$\hat{\psi_1} = logit(\beta_0)$$

For $t$ > 1

$$\hat{\psi_t} = \hat{\psi}_{t-1}\phi_{t-1} + (1-\hat{\psi}_{t-1})\gamma_{t-1}$$


```{r, echo = T}
psi <- array(NA, nyears)
psi[1] <- psi1
for(t in 2:nyears){
  psi[t] <- psi[t-1]*phi + (1-psi[t-1])*gamma
}
psi
```

---
# Expected vs Realized Occupancy

```{r, fig.height = 5.5}
gg_1 <- data.frame(Year = 1:nyears,
                   Occupancy = c(psi,colSums(z)/nsites, colSums(y > 0)/nsites),
                   Type = rep(c('Expected', 'Realized', 'Observed'), each = nyears))
ggplot(gg_1, aes(x = Year, y = Occupancy, col = Type))+
  geom_line()+
  geom_point(cex = 5)+
  theme_classic()+
  theme(axis.text = element_text(size = 15), axis.title = element_text(size = 20), legend.text = element_text(size = 15), legend.title = element_text(size = 20))
```

---
## Nimble Model

```{r, eval = T, echo =T}
occ.mod <- nimbleCode({
for(i in 1:nsites){
  z[i,1] ~dbern(psi)
  y[i,1] ~dbinom(p, z[i,1]*nvisits)
  
  for(t in 2:nyears){
    z[i, t] ~dbern(z[i,t-1]*(1-phi)+(1-z[i,t-1])*gamma)
    y[i, t] ~dbinom(p, z[i,t]*nvisits)
  } #end t
} #end i
  
gamma ~ dbeta(1, 1) 
phi ~ dbeta(1, 1) 
p ~ dbeta(1, 1)
psi ~ dbeta(1,1)
})
```


---
## Nimble Model

```{r, eval = T, echo =T}
params <- c('gamma', 'phi', 'p', 'psi')
n.c <- list(nsites = nsites, nvisits = nvisits, nyears = nyears)
n.d <- list(y =y)
n.i <- function()list(p = runif(1), gamma = runif(1), phi = runif(1), psi = runif(1))
```

--

Initial values for $z$ can be tricky - we need to make sure we don't suggest a site is unoccupied if we had observations there 
<br/>
--
One option is to start anywhere with y = 0 as unoccupied

```{r, eval = T, echo =T}
z.init <- (y >0)*1
n.i <- function()list(p = runif(1), gamma = runif(1), phi = runif(1), psi = runif(1), z.init =z.init)
```

--
Alternatively, flip a coin for any sites where y = 0
```{r, eval = T, echo =T}
z.init <- matrix(rbinom(nsites*nyears, 1, .5), ncol = nyears)
z.init[which(y > 0)] <- 1 
n.i <- function()list(p = runif(1), gamma = runif(1), phi = runif(1), psi = runif(1), z =z.init)
```

---
## Run the model in Nimble

```{r, echo = T, eval = F}
occ_out <- nimbleMCMC(code = occ.mod,
                     data = n.d,
                     constants = n.c,
                     inits = n.i(),
                     monitors = params,
                     thin = 1,
                     niter = 5000,
                     nburnin = 2500,
                     nchains = 3,
                     samplesAsCodaMCMC = TRUE
                      )
```

```{r}
occ_out <- readRDS('occ_out1.rds')
```
---
## Check the model
```{r, eval=T, fig.height=4.5, fig.width=8}
MCMCvis::MCMCtrace(occ_out, pdf = F)
```

---
## Compare with simulation
```{r, eval=T}
est <- (summary(occ_out)$quantiles)
est <- cbind(est, c(gamma, p, phi, psi1))
colnames(est)[6] <- 'true'
knitr::kable(est)
```

---
## Expansions

Royle and Dorazio (2008) suggested a reparameterization of this model that can sometimes help with convergence. 

<br/> 

Before we had: 
$$\Large z_{i,t} \sim Bernoulli(z_{i,t-1}(1-\phi)+(1-z_{i,t-1})\gamma)$$
<br/>

--
Instead, let:

$$\Large z_{i,t} \sim Bernoulli(\pi_{i,t})$$

$$\Large logit(\pi_{i,t}) = \alpha_{t-1} + \beta_{t-1}*z_{i,t-1}$$

---
## Expansions

$$\Large logit(\pi_{i,t}) = \alpha_{t-1} + \beta_{t-1}*z_{i,t-1}$$

We can then derive $\phi$ and $\gamma$

$$\Large logit(\phi_{i,t}) = \alpha_{t-1} + \beta_{t-1}$$

$$\Large logit(\gamma_{i,t}) = \alpha_{t-1}$$

