---
title: "Lecture 14"
subtitle: "Closed Spatial Capture Recapture"
author: "<br/><br/><br/>Spring 2025"
output:
  xaringan::moon_reader:
    css: ["default", "FANR8370.css", "FANR8370-fonts.css"]
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
      ratio: '16:9'
      slideNumberFormat: '%current%' 
---

```{r setup, include = FALSE}
options(htmltools.dir.version = FALSE)
knitr::opts_chunk$set(echo = FALSE, fig.align = 'center', warning=FALSE, message=FALSE,fig.retina = 2)
library(WILD8370)
library(nimble)
library(MCMCvis)
library(ggforce)
source(here::here("R/zzz.R"))
library(rphylopic) 
library(ggforce)
library(terra)
# library(gganimate)
```
## Readings

> 

---
## Motivations for SCR

With non-spatial capture recapture, we were able to estimate abundance

<br/> 
Unbiased estimates of $\large N$ require *accurate* estimates of $\large p$

--

<br/> 

Non-spatial models don't account for several important sources of variation in $p$  


  + Distance between animals and traps 
  
  + Trap-specific covariates 
  
  + Spatial variation in habitat or resources within the trapping array 

--

<br/> 
SCR makes it possible to estimate **density** not just $N$ in an unknown region 

---
## Motivations for SCR

With spatial data, we can ask new quetions

  + What influences spatial variation in density?
  
  + How do survival and recruitment vary in space and time?
  
  + How does movement influence density and detectability?

--

<br/>

Rather than think of SCR as a brand new estimation tool, you can think of it as a hierarchical framework combines different aspects of the environment to improve inference on population dynamics

---
## Homogenous SCR Model

Imagine a bunch of penguins on an iceberg. There is nothing on the iceberg but penguins and ice.

<br/>
```{r, fig.width= 6, fig.height= 6}
set.seed(1)
s <- data.frame(x = runif(10, -1, 1), y = runif(10, -1, 1))
s2 <- s+.025
s2$y <- s2$y+.05
s2$ind <- as.character(1:10)
s2$color <- rainbow(10)
s$ind <- s2$ind
s$color <- s2$color
uuid <- get_uuid( 'Eudyptes chrysolophus', n = 2)
img_svg <- get_phylopic(uuid[2], format = "vector")
ggplot(s, aes(x =x, y= y))+
  add_phylopic(img_svg, alpha=0.9, x= s2$x, y=s2$y, height = .2, col = s2$color)+
  xlim(-1.5,1.5)+
  ylim(-1.5, 1.5)
```

---
## Homogenous SCR Model

The penguins are cold, so they shuffle around a bit, but they generally stay somewhat near their original positions, where they have left their children (eggs).

<br/>
```{r, fig.width= 6, fig.height= 6}
ggplot(s, aes(x =x, y= y))+
  geom_circle(dat = s, aes(x0 = x, y0 = y, r = .35), lty = 2, alpha = .15)+
  add_phylopic(img_svg, alpha=0.9, x= s2$x, y=s2$y, height = .2, col = s2$color)+
  xlim(-1.5,1.5)+
  ylim(-1.5, 1.5)+
  scale_fill_manual(values = rainbow(10))+
  #scale_color_manual(values = rainbow(10))+
  theme(legend.position = 'none')
```

---
## Homogenous SCR Model

I wish to calculate the abundance of penguins, so I set out 25 fish traps to catch them.

<br/>
```{r, fig.width= 6, fig.height= 6}
x <- seq(-.3, 1, .3)
traps <- expand.grid(x,x)
uuid2 <- '381e7e19-bb44-4432-85ce-627f05d159a3'
img2 <- get_phylopic(uuid2, format = "vector")
ggplot(s, aes(x =x, y= y))+
  geom_circle(aes(x0 = x, y0 = y, r = .35), lty = 2, alpha = .15, lwd = .2)+
  add_phylopic(img_svg, alpha=0.9, x= s2$x, y=s2$y, height = .2, col = s2$color)+
  add_phylopic(img2, alpha = .9, x = traps$Var1, y = traps$Var2, height = .05, col = 'blue')+
  xlim(-1.5,1.5)+
  ylim(-1.5, 1.5)
```

---
## Homogenous SCR Model

Penguins that live closer to the traps will be capture more often than those that don't.

<br/>
```{r, fig.width= 6, fig.height= 6}
g0 <- .7
dist <- p <- y <- array(NA, c(10, 25))
sigma <- .35/2
for(j in 1:25) {
  for(i in 1:10){
    dist[i,j] <- sqrt((s$x[i]-traps$Var1[j])^2 + (s$y[i]-traps$Var2[j])^2)
    p[i,j] <- g0*exp(-dist[i,j]^2/(2*sigma^2))
    y[i,j] <- rbinom(1, 3, p[i,j]) ## Model for the capture histories
  }
}

trap_segs <- data.frame(x0 = 1, y0 = 2, x1 = 3, y1= 4, col = NA)
for(i in 1:10){
  mytraps <- which(y[i,] > 0)
  if(length(mytraps) == 0) next 
  for(k in 1:length(mytraps)){
    add <- data.frame(x0 = traps[mytraps[k],1], y0 = traps[mytraps[k],2], 
                      x1 = s[i,1], y1 = s[i, 2], col = s[i,3])
    trap_segs <- rbind(trap_segs, add)
  }
}
trap_segs <- trap_segs[-1,]

ggplot(s, aes(x =x, y= y))+
  geom_circle(dat = s, aes(x0 = x, y0 = y, r = .35), lty = 2, lwd = .2)+
  add_phylopic(img_svg, alpha=0.9, x= s2$x, y=s2$y, height = .2, col = s2$color)+
  add_phylopic(img2, alpha = .9, x = traps$Var1, y = traps$Var2, height = .05, col = 'blue')+
  geom_segment(dat = trap_segs, aes(x = x0, xend = x1, y = y0, yend= y1, col = col), lty = 2)+
  xlim(-1.5,1.5)+
  ylim(-1.5, 1.5)+
  scale_color_manual(values = list('1' = s$color[1], '2' = s$color[2],'3' = s$color[3],'4' = s$color[4],
                                   '5' = s$color[5], '6' = s$color[6], '7' = s$color[7], '8' = s$color[8],
                                   '9' = s$color[9], '10' = s$color[10]))+
  theme(legend.position = 'none')
```

---
## SCR Model

When we model spatial data, it would be convenient if two things were true:

First:


 $$ N_t \sim Pois(\lambda_t)$$
$$ log(\lambda_t) = \beta0 + \beta1*x + \cdots$$
This formulation is familiar and lets us put covariates on $\lambda$ that tell us the expected abundance at some known resolution (per pixel, per site, etc.). 

--
<br/>
But we also want to use individual level data, such that:

$$ z_i \sim Bernoulli(\psi_t)$$ 
$$E(N_t) = \psi_t*M$$

$$N_t = \sum_{1}^{M} z_i$$
*where $M >> N$
---
## SCR Model

This is inconvenient, because to use MCMC to solve for parameters, you can't have two different ways to define the same variable. 

--

Luckily, we can do some rearranging to make things work nicely. 


If: 
$$E(N_t) = \psi_t*M$$
and 

$$E(N_t) = \lambda_t$$

Then:

$$\psi_t = \frac{E(N_t)}{M} = \frac{\lambda_t}{M}$$
--

If space is homogenous, we can just replace this with:

$$\psi_t \sim Uniform(0, 1)$$


---
## Homogeneous SCR Model

If space is homogenous, we don't have to do much else. We can just add in a quick prior for locations:

$$\large \mathbf{s_{i}} \sim Uniform(s_{min}, s_{max})$$
Note that *ALL* animals receive an activity center location regardless if they are real or not. We just can't detect the fake ones. 

--

The observation process is then just based on distance between the activity center and the trap in question. 

$$\large p_{ij} = g_0*e^{(-\|s_i - x_j\|^2/(2\sigma^2))}$$

$$\large y_{ijk} \sim Bernoulli(z_i*p_{ij})$$
** Note: this observation process assumes multiple animals can be captured in multiple traps on the same occasion. It is still often used for single-catch traps (like mammal traps) simply because the alternatives require knowledge of capture time. 

---
## Homogeneous SCR Model Simulation

Let's simulate a simple closed, homogeneous SCR model using the following parameters:

  + `set.seed(15)`
  
  For the real individuals:
  
    +  N = 25
  
    +  locations bound between [0, 1] in both directions
  
  For the observation process:
  
    +  $g_0$ = .8 
    
    +  $\sigma$ = .5
    
    +  traps located every .2 units from (.2, .2) to (.8, .8) 
    
        + Hint: `x <- seq(.2, .8, .2); expand.grid(x,x)`
        
    +  3 occasions (trials) 

---
## Homogeneous SCR Model Simulation

```{r, echo = T}
set.seed(15)
N <- 25
max <- 1
min <- 0
g0 <- .8
sig <- .5
x <- seq(.2, .8, .2)
traps <- expand.grid(x,x)
ntrials <- 3
ntraps <- nrow(traps)
s <- matrix(runif(50), ncol = 2)
dist <- p <- y <- array(NA, c(N, ntraps))
for(j in 1:ntraps) {
  for(i in 1:N){
    dist[i,j] <- sqrt((s[i,1]-traps[j,1])^2 + (s[i,2]-traps[j,2])^2)
    p[i,j] <- g0*exp(-dist[i,j]^2/(2*sig^2))
    y[i,j] <- rbinom(1, ntrials, p[i,j]) ## Model for the capture histories
  }
}
```

---
## Homogeneous SCR Model in Nimble

```{r, echo = T}
nimbleCode({
g0 ~ dunif(0, 1)       ## Baseline capture probability
sigma ~ dunif(0, 0.5)  ## Scale parameter of encounter function
psi ~ dunif(0, 1)      ## Data augmentation parameter

for(i in 1:M) {
  s[i,1] ~ dunif(xlim[1], xlim[2]) ## x-coord of activity center
  s[i,2] ~ dunif(ylim[1], ylim[2]) ## y-coord of activity center
  z[i] ~ dbern(psi)                ## Is individual real?
  for(j in 1:J) {
    dist[i,j] <- sqrt((s[i,1]-x[j,1])^2 + (s[i,2]-x[j,2])^2)
    p[i,j] <- g0*exp(-dist[i,j]^2/(2*sigma^2))
    y[i,j] ~ dbinom(p = z[i]*p[i,j], size = ntrials) ## Model for the capture histories
    }
  }

EN <- M*psi  ## Expected value of N
N <- sum(z)  ## Realized value of N
}
)
```


---
## Inhomogeneous SCR Model

The previous model is great if we don't have any spatial covariates. We don't even have to worry about $\lambda$ at all, we just slap a uniform prior on $\psi$ and we're good to go. 

<br/>

But when we *DO* have spatial variation, we are left with an issue. 

<br/>

How do we ensure that the locations of our $M$ individuals (real or not) follow the pattern of $Pois(\lambda_t)$? 


$$s_i \propto \lambda_t(s)$$ 
--

In continuous space, we would say:

$$\Lambda = \int_{\mathcal{S}} \lambda(s)$$
In english, this says that the area under the curve for any given region is the expected number of points (individuals) in that region. 

---
## Inhomogeneous SCR Model

In discrete space, this is similar to creating a raster where the pixel values are equal to the number of individuals in each pixel and then randomly spreading those individuals throughout that pixel. 

<br/>

As the pixel size gets smaller and smaller, we start to approximate continuous space. 


<br/> 

This is called a *spatial point process*. The equation ($\lambda(s)$) describes spatial variation in the density of points. We call this an *intensity function*. 

<br/>

Our goal here is to estimate the intensity function so that we can estimate how many individuals are located in each pixel in our area of interest. 

---
## Inhomogeneous SCR Model

```{r, fig.width= 6, fig.align='center'}
set.seed(55)
par(oma = c(0,0,0,0))
delta <- 0.25/6 ## Resolution
grid0 <- seq(delta/2, 1-delta/2, delta)
grid <- expand.grid(grid0, grid0)
distmat <- as.matrix(dist(grid))
npixels <- nrow(distmat)
## Covariance matrix
V <- exp(-0.4*distmat)
R <- chol(V)
X <- t(R) %*% rnorm(npixels)

Canopy <- rast(nrows = length(grid0), ncols = length(grid0), extent = c(0, 1, 0, 1))
Canopy[] <- X[,1]
plot(Canopy) 
```
Consider the above raster of canopy cover. 

---
## Inhomogeneous SCR Model

What if I thought the relationship betwen abundance and canopy cover was a simple poisson regression? 

$$ N \sim Pois(\lambda) $$
$$log(\lambda) = \beta_0 + \beta_1*canopy$$

```{r, fig.width= 6, fig.height= 6}
X2 <- exp(.5 + 2*X[,1])
Canopy[] <- X2
plot(Canopy) 
```

---
## Inhomogeneous SCR Model

If we normalize this raster, we get the relative probability of an individual being in that pixel vs any other pixel. This is the same as $\frac{\lambda}{\sum\lambda}$, so we don't *need* the raster.

```{r, fig.width= 6, fig.height= 6}
X2 <- exp(.5 + 2*X[,1])
Canopy[] <- X2/sum(X2)
plot(Canopy) 
```

---
## Inhomogeneous SCR Model

The problem is that this probability layer is not a distribution. So how do we trick Nimble or any other MCMC software into cooperating? 

<br/>

The secret is a fun trick of the poisson distribution! 

--

<br/>

Remember that:

$$P(Pois(\lambda) = 0) = e^{-\lambda}$$

So we fit this model:

$$0 \sim Pois(-logProb[i])$$
For any given pixel $k$ and individual $i$, the log probability of that individual being located in that pixel is:

$$logProb[i] = log(\frac{\lambda[i]}{sum(\lambda)})$$


---
## Inhomogeneous SCR Model

This means we're saying:

$$\lambda = -logProb[i]$$
Therefore:

$$P(Pois(\lambda) = 0) = e^{-\lambda} = e^{-(-logProb[i])} = e^{logProb[i]} $$ 

Remember that `log(1) = 0`. 

This means that as the probability of individual $i$ being in pixel $k$ approaches 1, `logProb[i]` approaches 0 and the likelihood of `Pois(-logProb[i]) = 0` also approaches 1. Nifty! 

---
## Inhomogeneous SCR Model in Nimble

In practice, this is accomplished fairly easily in nimble, though it adds a lot of lines to our code which is annoying.

```{css, echo = F}
.smaller .remark-code { 
font-size: 45% !important;
}
```

.smaller[
```{r, echo = T}
ipp <- nimbleCode({

psi ~ dunif(0, 1)  ## Can't put prior on psi *and* beta0
beta0 <- log(M*psi/SumLambda) ## Algebra
beta1 ~ dnorm(0, 0.1)
g0 ~ dunif(0, 1)
sigma ~ dunif(0, 0.5)

for(g in 1:G) { ## Loop over pixels
  lambda[g] <- exp(beta1*canopy[g])*pixelArea
  pi[g] <- lambda[g]/SumLambda
}
SumLambda <- sum(lambda)

for(i in 1:M) {
  s[i,1] ~ dunif(xlim[1], xlim[2]) 
  s[i,2] ~ dunif(ylim[1], ylim[2])
  pixel[i] <- lookup[round((ylim[2]-s[i,2])/delta+0.5),  ## row
                     round((s[i,1]-xlim[1])/delta+0.5)]  ## column
  logProb[i] <- log(pi[pixel[i]])
  zeros[i] ~ dpois(-logProb[i]) ## zeros trick for IPP
  z[i] ~ dbern(psi)
  for(j in 1:J) {
    dist[i,j] <- sqrt((s[i,1]-x[j,1])^2 + (s[i,2]-x[j,2])^2)
    p[i,j] <- g0*exp(-dist[i,j]^2/(2*sigma^2))
  }
}
for(i in 1:M) {  ## Model for capture histories
  for(j in 1:J) {
    y[i,j] ~ dbinom(prob = p[i,j]*z[i], size = K)
  }
}

EN <- M*psi
N <- sum(z)
})
```
]

---
## Inhomogeneous SCR Model in Nimble

Let's go over this line by line. 

<br/>
First, the easy part - Detection! We can only see animals that are real. No false positives! 

```{r, echo = T, eval = F}
for(i in 1:M) { 
  for(j in 1:J) {
    y[i,j] ~ dbinom(prob = p[i,j]*z[i], size = K)
  }
}
```

---
## Inhomogeneous SCR Model in Nimble

Our probability of detection is based on the distance between our trap and the animal's activity center. 

```{r, echo = T, eval = F}
z[i] ~ dbern(psi)
  for(j in 1:J) {
    dist[i,j] <- sqrt((s[i,1]-x[j,1])^2 + (s[i,2]-x[j,2])^2)
    p[i,j] <- g0*exp(-dist[i,j]^2/(2*sigma^2))
```


---
## Inhomogeneous SCR Model in Nimble

Individuals can be anywhere in the state space. We can figure out which pixel they are in by using a lookup grid that tells us the pixel number. If you were to number pixels on a raster, the first one is the top left corner, the last one is the bottom right corner. Delta is the pixel width 

```{r, echo = T, eval = F}
s[i,1] ~ dunif(xlim[1], xlim[2]) 
s[i,2] ~ dunif(ylim[1], ylim[2])
pixel[i] <- lookup[round((ylim[2]-s[i,2])/delta+0.5),  ## row
                    round((s[i,1]-xlim[1])/delta+0.5)]  ## column
```

---
## Inhomogeneous SCR Model in Nimble

Finally (ignoring priors), these parts are our zeros trick

```{r, echo = T, eval = F}
logProb[i] <- log(pi[pixel[i]])
zeros[i] ~ dpois(-logProb[i]) ## zeros trick for IPP
```

```{r, echo = T, eval = F}
for(g in 1:G) { ## Loop over pixels
  lambda[g] <- exp(beta1*canopy[g])*pixelArea
  pi[g] <- lambda[g]/SumLambda
}
SumLambda <- sum(lambda)
```


---
## Inhomogeneous SCR Model in Nimble

In lab we will explore how to run these models in Nimble and learn a speed trick to make our model run faster!

