---
title: "lecture 5: generalized linear model review"
subtitle: "FANR 8370"
author: "<br/><br/><br/>Spring 2025"
output:
  xaringan::moon_reader:
    css: ["default", "FANR8370.css", "FANR8370-fonts.css"]
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
      ratio: '16:9'
      slideNumberFormat: '%current%' 
---

```{r setup, include = FALSE}
options(htmltools.dir.version = FALSE)
knitr::opts_chunk$set(echo = FALSE, fig.align = 'center', warning=FALSE, message=FALSE,fig.retina = 2)
library(WILD8370)
source(here::here("R/zzz.R"))
library(nimble)
# library(gganimate)
```

# readings

> ### Kéry & Schaub 48-55

---
class: inverse, center, middle

# from linear models to glm

---
# linear models

<br/>
<br/>
$$\Large response = deterministic\; part+stochastic\; part$$ 
<br/>
<br/>

--

$$\underbrace{\LARGE \mu_i = \beta_0 + \beta_1 \times x_i}_{Deterministic}$$
<br/>
<br/>

--

$$\underbrace{\LARGE y_i \sim normal(\mu_i, \tau)}_{Stochastic}$$  

???

Note that the deterministic portion of the model has the same form as the equation for a line: $y = a + b \times x$, which is why we call these linear models

---
class: inverse, middle, center

# linear models under the hood

## Variations on the stochastic model

---
# stochasticity in the linear model

$$\underbrace{\LARGE \mu_i = -2 + 0.5 \times x_i}_{Deterministic}$$
--

```{r fig.height=4, fig.width=4}
x <- rbinom(100, size = 1, prob = 0.5)
y <- -2 + 0.5 * x + rnorm(100, 0, 0.25)

df <- data.frame(x = x, y = y)

ggplot(df, aes(x, y)) + geom_boxplot(color = "white") +
  geom_segment(aes(x = -0.1, xend = 0.1, y = -2, yend = -2), color = WILD8370_colors$value[WILD8370_colors$name == "primary"], size = 1.5) +
  geom_segment(aes(x = 0.9, xend = 1.1, y = -1.5, yend = -1.5), color = WILD8370_colors$value[WILD8370_colors$name == "primary"], size = 1.5) +
  scale_y_continuous(expression(mu))
```

---
# stochasticity in the linear model

$$\LARGE \mu_i = -2 + 0.5 \times x_i$$

$$\underbrace{\LARGE y_i \sim Normal(\mu_i, \tau)}_{Stochastic}$$  

--

```{r fig.height=4, fig.width=4}
ggplot(df, aes(x, y, group = x)) + 
  geom_boxplot(width = 0.25) + 
    geom_segment(aes(x = -0.1, xend = 0.1, y = -2, yend = -2), color = WILD8370_colors$value[WILD8370_colors$name == "primary"], size = 1.5) +
  geom_segment(aes(x = 0.9, xend = 1.1, y = -1.5, yend = -1.5), color = WILD8370_colors$value[WILD8370_colors$name == "primary"], size = 1.5) +
  geom_point(color = WILD8370_colors$value[WILD8370_colors$name == "warning"], alpha = 0.25)
```

---
# stochasticity in the linear model

$$\LARGE \mu_i = -2 + 0.5 \times x_i$$


```{r fig.height=4, fig.width=4}
x <- rnorm(100)
y <- -2 + 0.5 * x + rnorm(100, 0, 0.25)

df <- data.frame(x = x, y = y)

ggplot(df, aes(x, y)) + geom_point(color = "white") +
  geom_abline(slope = 0.5, intercept = -2) +
  scale_y_continuous(expression(mu))
```

---
# stochasticity in the linear model

$$\LARGE \mu_i = -2 + 0.5 \times x_i$$

$$\LARGE y_i \sim Normal(\mu_i, \tau)$$  

```{r fig.height=4, fig.width=4}
ggplot(df, aes(x, y)) +
  geom_abline(slope = 0.5, intercept = -2) +
  geom_point(color = WILD8370_colors$value[WILD8370_colors$name == "warning"], alpha = 0.25)
```

---
class: inverse, center, middle

# components of the linear model

---
# components of the linear model

### 1) Distribution

$$\LARGE y_i \sim normal(\mu_i, \tau)$$

--

### 2) Link function

$$\LARGE \mu_i = E(y_i) = linear\;\;predictor$$

--

### 3) Linear predictor

$$\LARGE \beta_0 + \beta_1 \times x_i$$


---
# stochasticity in the linear model

#### What happens if $\Large 0 \leq y_i$?
<br/>
<br/>

--

```{r fig.height=4, fig.width=4}
ggplot(df, aes(x, y)) +
  geom_point(color = "white") +
  geom_abline(slope = 0.5, intercept = -2) +
  scale_y_continuous(expression(mu))

```

---
#  components of the generalized linear model

#### 1) Distribution

$$\LARGE y_i \sim normal(\mu_i, \tau)$$

---
# components of the generalized linear model

#### 1) Distribution

$$\LARGE y_i \sim Poisson(\lambda_i)$$


--

#### 2) Link function

$$\LARGE \lambda_i = E(y_i) = linear\;\;predictor$$

---
# components of the generalized linear model

#### 1) Distribution

$$\LARGE y_i \sim Poisson(\lambda_i)$$


#### 2) Link function

$$\LARGE log(\lambda_i) = log(E(y_i)) = linear\;\;predictor$$

--
```{r fig.height = 2.5, fig.width=5}
lambda <- runif(100, 0, 20)
l.lambda <- log(lambda)

df <- data.frame(lambda = lambda, log.lambda = l.lambda)

ggplot(df, aes(x = log.lambda, y = lambda)) + geom_line(size = 2, color = WILD8370_colors$value[WILD8370_colors$name=="warning"]) +
  scale_y_continuous(name = expression(lambda)) +
  scale_x_continuous(name = expression(paste("log(", lambda, ")")))
```

---
# components of the generalized linear model

#### 1) Distribution

$$\LARGE y_i \sim Poisson(\lambda_i)$$


#### 2) Link function

$$\LARGE log(\lambda_i) = log(E(y_i)) = linear\;\;predictor$$

---
# components of the generalized linear model

#### 1) Distribution

$$\LARGE y_i \sim Poisson(\lambda_i)$$


#### 2) Link function

$$\LARGE log(\lambda_i) = log(E(y_i)) = linear\;\;predictor$$

#### 3) Linear predictor

$$\LARGE \beta_0 + \beta_1 \times x_i$$ 


---
# components of the generalized linear model

#### 1) Distribution

$$\LARGE y_i \sim Bernoulli(p_i)$$

--

#### 2) Link function

$$\LARGE logit(p_i) = log \bigg(\frac{p_i}{1-p_0}\bigg) = linear\;\;predictor$$

--

```{r fig.height = 2.5, fig.width=5}
p <- runif(100)
l.p <- log(p/(1-p))

df <- data.frame(p = p, logit.p = l.p)

ggplot(df, aes(x = logit.p, y = p)) + geom_line(size = 2, color = WILD8370_colors$value[WILD8370_colors$name=="warning"]) +
  scale_x_continuous(name = "logit(p)")
```

---
# components of the generalized linear model

#### 1) Distribution

$$\LARGE y_i \sim Bernoulli(p_i)$$

#### 2) Link function

$$\LARGE logit(p_i) = log \bigg(\frac{p_i}{1-p_0}\bigg) = linear\;\;predictor$$

#### 3) Linear predictor

$$\LARGE \beta_0 + \beta_1 \times x_i$$ 

---
# components of the generalized linear model

#### 1) Distribution

$$\LARGE y_i \sim binomial(N, p_i)$$

#### 2) Link function

$$\LARGE logit(p_i) = log \bigg(\frac{p_i}{1-p_0}\bigg) = linear\;\;predictor$$

#### 3) Linear predictor

$$\LARGE \beta_0 + \beta_1 \times x_i$$ 

---
# glms in nimble - normal link

$$\mu_i = \beta_0 + \beta_1 \times x_i$$
$$y_i \sim Normal(\mu_i, \tau)$$  

```{r, eval=F, echo = T}
nimbleCode({
  ## Priors
    beta0 ~ dnorm(0, sd = 10) #mean, sd
    beta1 ~ dnorm(0, sd = 10)
    sigma ~ dgamma(.001,.001) # residual sd
    
  ## Likelihood
    for(i in 1:N){
      mu[i] <- beta0 + beta1 * x[i]
      y[i] ~ dnorm(mu[i], sd = sigma)
    }
}) 
```

---
# glms in nimble - poisson Link

$$y_i \sim Poisson(\lambda_i)$$

$$log(\lambda_i) = log(E(y_i)) = \beta_0 + \beta_1 \times x_i$$

```{r, eval=F, echo = T}
nimbleCode({
  ## Priors
    beta0 ~ dnorm(0, sd = 10) #mean, precision
    beta1 ~ dnorm(0, sd = 10)
    
  ## Likelihood
    for(i in 1:N){
      log(lambda[i]) <- beta0 + beta1 * x[i]
      #Equivalently:
      #lambda[i] <- exp(beta0 + beta1 * x[i])
      y[i] ~ dpois(lambda[i])
    }
}) 
```

---
# glms in nimble - binomial link

$$y_i \sim binomial(N, p_i)$$

$$logit(p_i) = log \bigg(\frac{p_i}{1-p_0}\bigg) = \beta_0 + \beta_1 \times x_i$$

```{r binomial, eval=F, echo = T}
nimbleCode({
  ## Priors
    beta0 ~ dnorm(0, sd = 10) #mean, sd
    beta1 ~ dnorm(0, sd = 10)
    
  ## Likelihood
    for(i in 1:nobs){
      logit(p[i]) <- beta0 + beta1 * x[i]
      #Equivalently:
      #p[i] <- exp(beta0 + beta1 * x[i])/(1+exp(beta0 + beta1 * x[i]))
      y[i] ~ dbinom(p[i], J) #J = trials
    }
}) 
```

---
class: inverse

# generalized linear models
<br/>
<br/>

--

- Flexible method to model observations arising from different probability distributions  
<br/>
<br/>

--

- Link many classical tests under unified framework  
<br/>
<br/>

--

- Underlie nearly all common ecological models  

